---
title: "kernelTDA - Vignette"
subtitle: "Statistical Learning with Kernels for Topological Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{kernelTDA-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Topological Data Analysis (TDA) is a relatively new branch of statistics devoted to the estimation of the connectivity structure of data, via means of topological invariants such as (Persistent) Homology Groups. Topology provides a highly interpretable characterization of data, hence TDA has seen an impressive growth in the last couple of years. 


While several inference-ready tools have been developed in the framework of TDA, their popularity The hype on the theoretical side however, has not been matched by the same popularity in applications. 
The `kernelTDA` package aims at filling this gap by providing --- tools, with a special emphasis on kernels. The main contribution of `kernelTDA` is in fact to provide an `R` implementation of the most popular kernels to be used in the space of Persistence Diagrams:

- Persistence Scale Space Kernel
- Gaussian Weighted Kernel
- Sliced Wasserstein Kernel
- Fisher Information Kerner
- Geodesic Kernels

In addition, it also contains an `R` version of the Persistence Images, and a port to the `C++` library `HERA`, which allows to compute any Wasserstein distance between Persistence Diagrams. 


## Topological Kernels 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(kernelTDA)
```
